---
title: "Reanalysis of a Data set"
author: "Kilian Sander, Yves Wycisk, Reinhard Kopiez"
date: '2022'
output: html_document
csl: inst/apa.csl
---
# Preparation
## Data, Packages, and Design
Load the data:
```{r dataset, messages=FALSE}
load("data/data_raw.Rda")
```
Load relevant packages:
```{r packages, message=FALSE, warning=FALSE, results='hide'}
sapply(c("tidyverse", "MANOVA.RM", "rstatix", "ggpubr"), require, character.only = T)
```
Prepare the design with vectors:
```{r formats and GEMS, message=FALSE}
formats <- c("stereo", "surround", "auro")
gems_factors <- c("wonder", "transcendence", "tenderness", "nostalgia",
                  "peacefulness", "power", "joyful_activation", "tension",
                  "sadness")
# item to factor mapping
wonder <- c("moved", "filled_with_wonder", "allured")
transcendence <- c("feeling_of_transcendence", "fascinated", "overwhelmed")
tenderness <- c("affectionate", "tender", "mellowed")
nostalgia <- c("sentimental", "dreamy", "nostalgic")
peacefulness <- c("calm", "serene", "soothed")
power <- c("energetic", "triumphant", "strong")
joyful_activation <- c("joyful", "animated", "bouncy")
tension <- c("agitated", "nervous", "tense")
sadness <- c("sad", "sorrowful", "tearful")
```
## Filtering
Exclude a participant if there are at least two `NA`s for a GEMS factor of one stimulus:
```{r filter cases, message=FALSE}
data_raw$complete <- TRUE
for (p in data_raw$participant) {
  part_data <- data_raw %>% filter(participant == p)
  for (excerpt in c("a", "b")) {
    for (format in formats) {
      for (factor in gems_factors) {
        temp_data <- part_data %>%
          dplyr::select(starts_with(paste(excerpt, format, get(factor), sep = "_")))
        nas <- temp_data %>% t() %>% is.na() %>% sum()
        if (nas > 1) {
          data_raw$complete[p] <- FALSE
        }
      }
    }
  }
}
rm(part_data, temp_data, nas, p, format, excerpt)
data_filtered <- data_raw %>% filter(complete==TRUE)
N <- length(data_filtered$participant)
```
After removing incomplete cases the data set contains `r N` participants.

## Scoring
Calculate mean values for each item per format:
```{r item means, message=FALSE, warning=FALSE}
for (format in formats) {
  for (factor in gems_factors) {
    for (item in get(factor)) {
      item_mean_vector <- data_filtered %>%
        select(ends_with(paste0(format, "_", item))) %>%
        rowMeans(na.rm = TRUE)
      eval(str2expression(paste0("data_filtered$",
                                 format, "_", item,
                                 " <- item_mean_vector")))
    }
  }
}
rm(item, item_mean_vector)
data_filtered <- data_filtered %>% mutate_all(~ifelse(is.nan(.), NA, .))
```
Calculate the mean of each factor per format across its items:
```{r factor means, message=FALSE, warning=FALSE}
for (format in formats) {
  for (factor in gems_factors) {
    factor_mean_vector <- data_filtered %>%
      select(paste0(format, "_", get(factor))) %>% 
      rowMeans(na.rm = TRUE)
    eval(str2expression(paste0("data_filtered$",
                                 format, "_", factor,
                                 " <- factor_mean_vector")))
  }
}
rm(format, factor, factor_mean_vector)
```
## Long format
```{r long formats, message=FALSE, warning=FALSE}
vars_vector <- sapply(formats,
                      function(format) {
                        paste0(format, "_", gems_factors)
                      },
                      USE.NAMES = F) %>% as.vector()
data_long_mult <- data_filtered %>% select(participant, vars_vector)
names(data_long_mult) <- names(data_long_mult) %>%
  gsub("joyful_activation", "joyfulactivation", .)
data_long_mult <- data_long_mult %>%
  pivot_longer(!participant,
               names_to = c("format", ".value"),
               names_pattern = "(.*)_(.*)")
data_long_single <- data_long_mult %>%
  pivot_longer(!c(participant, format),
               values_to = "score",
               names_to = "factor") %>% 
  mutate(factor = recode(factor,
                         joyfulactivation = "joyful activation") %>% tools::toTitleCase(),
         format = format %>% tools::toTitleCase() %>% as_factor())
names(data_long_mult) <- names(data_long_mult) %>%
  gsub("joyfulactivation", "joyful_activation", .)
data_long_mult$format <- data_long_mult$format %>% as_factor()
```
# Analyses
## Descriptives
```{r age summary, include=FALSE}
age_summary <- summary(data_filtered$age)
```
Out of the `r N` participants `r table(data_filtered$gender)[["female"]]`
indicated "female" as their gender whereas `r table(data_filtered$gender)[["male"]]`
indicated "male".
The participants' age ranged from `r age_summary[[1]]` to `r age_summary[[6]]`
with a mean of `r age_summary[[4]]` (*SD* = `r sd(data_filtered$age)`).

```{r descriptives, echo=FALSE, message=FALSE, warning=FALSE}
descriptives <- data_long_single %>% group_by(factor, format) %>%
  summarise(M = mean(score), SD = sd(score))
descriptives %>% 
  knitr::kable(col.names = c("GEMS Factor", "Format", "$M$", "$SD$"),
               align = "llcc")
```
```{r mean plot, echo=FALSE, fig.width=10, message=FALSE, warning=FALSE}
data_long_single %>% group_by(format, factor) %>%
  summarise(M = mean(score), SD = sd(score)) %>%
  ggplot(., aes(x = factor, y = M, color = format)) +
  geom_point(aes(shape = format)) +
  geom_path(aes(group = format, linetype = format)) +
  ylim(1,5) +
  theme_pubr() +
  theme(legend.position = c(.99,.99),
        legend.justification = c(1,1)
        ) +
  xlab("GEMS Factor") +
  ylab("Mean") +
  labs(color = "Format", shape = "Format", linetype = "Format")
```

## Multivariate Analysis of Variance

```{r manova}
#rm_manova <-
#  multRM(formula = cbind(wonder, transcendence, tenderness,
#                         nostalgia, peacefulness, power,
#                         joyful_activation, tension, sadness) ~ format,
#         data = data_long_mult,
#         subject = "participant",
#         within = "format",
#         para = TRUE,
#         iter = 50000,
#         dec = 5,
#         seed = 987)
load(file = "results/rm_manova.Rda")
summary(rm_manova)
```
## Analysis of Variance for *Transcendence*
### Assumption Checks
#### Extreme Outliers
```{r anova outliers}
data_long_mult %>% group_by(format) %>%
  select(format, participant, transcendence) %>%
  identify_outliers(transcendence) %>% 
  filter(is.outlier)
ggboxplot(data_long_mult, x = "format", y = "transcendence", add = "point")
```

#### Normality
```{r anova normality}
data_long_mult %>% group_by(format) %>%
  shapiro_test(transcendence)
ggqqplot(data_long_mult, "transcendence", facet.by = "format")
```

#### Sphericity
```{r anova}
transcendence_anova <-
  anova_test(data = data_long_mult,
             dv = transcendence,
             wid = participant,
             within = format,
             type = 3,
             effect.size = "pes")
transcendence_ges <- anova_test(data = data_long_mult,
           dv = transcendence,
           wid = participant,
           within = format,
           type = 3,
           effect.size = "ges")$ANOVA$ges
transcendence_anova$`Mauchly's Test for Sphericity`
```
### Results
```{r anova results, echo=FALSE}
transcendence_res <- get_anova_table(transcendence_anova)
class(transcendence_res) <- "data.frame"
transcendence_res %>% mutate(ges = transcendence_ges) %>%
  select(!`p<.05`) %>% knitr::kable(col.names = c("Effect", "DFn", "DFd",
                                                  "$F$", "$p$",
                                                  "$\\eta^2_p$",
                                                  "$\\eta^2_G$"))
```

![](results/rm_anova_transcendence_plot.svg)

*Note.* Error bars represent 95% confidence intervals.
GEMS factors are on a 5-point Likert scale coded from 1 to 5.


|Contrasts         |Estimate|CI LL  |CI UL  |SE     |df   |$t$  |$p$     |
|:-----------------|:------:|:-----:|:-----:|:-----:|:---:|:---:|:------:|
|Stereo -- Surround|0.183   |0.023  |0.343  |0.081  |102  |2.266|0.026   |
|Stereo -- Auro    |0.413   |0.254  |0.573  |0.081  |102  |5.128|1.397e-6|
|Surround --Auro   |0.231   |0.071  |0.391  |0.081  |102  |2.862|0.005   |

## Correlations for *Transcendence*
### Correlations Between the Formats
```{r correlations, include=FALSE}
cor_stereo_surround <-
  cor.test(data_filtered$stereo_transcendence,
           data_filtered$surround_transcendence,
           alternative = "greater",
           method = "pearson",
           conf.level = .95)
cor_stereo_auro <-
  cor.test(data_filtered$stereo_transcendence,
           data_filtered$auro_transcendence,
           alternative = "greater",
           method = "pearson",
           conf.level = .95)
cor_surround_auro <-
  cor.test(data_filtered$surround_transcendence,
           data_filtered$auro_transcendence,
           alternative = "greater",
           method = "pearson",
           conf.level = .95)
cor_table <- data.frame(Formats = c("Stereo -- Surround",
                                    "Stereo -- Auro",
                                    "Surround -- Auro"),
                        r = c(cor_stereo_surround$estimate,
                              cor_stereo_auro$estimate,
                              cor_surround_auro$estimate),
                        p = c(cor_stereo_surround$p.value,
                              cor_stereo_auro$p.value,
                              cor_surround_auro$p.value),
                        LL = c(cor_stereo_surround$conf.int[1],
                              cor_stereo_auro$conf.int[1],
                              cor_surround_auro$conf.int[1]),
                        UL = c(cor_stereo_surround$conf.int[2],
                              cor_stereo_auro$conf.int[2],
                              cor_surround_auro$conf.int[2])) %>% 
  mutate(`Fisher's z` = atanh(r),
         p = sapply(p,
                    function(x) if(x < 10^-3) "< .001"))
```
```{r correlations table, echo=FALSE}
knitr::kable(cor_table %>% 
  mutate(across(where(is.numeric), round, digits = 4)),
             col.names = c("Formats", "$r$", "$p$", "CI LL", "CI UL", "Fisher's $z$"),
             align = "lccccc")
```

### Average Correlation
To get a mean correlation individual correlation coefficients have to be
Fisher $z$-transformed.
The Fisher $z$-transformation is equivalent to the inverse hyperbolic tangent.
```{r average correlations}
avg_cor_z <- mean(cor_table$`Fisher's z`)
avg_cor_r <- tanh(avg_cor_z)
```
Taking the averaged Fisher $z$-transformed values results in a value of
`r round(avg_cor_z, 4)`.
Back-transformed to Person's $r$ the average correlation between the three
audio formats for the GEMS factor Transcendence is `r round(avg_cor_r, 4)`.

## Common Language Effect Sizes for *Transcendence*
The common language effect size (CLE) expresses "the probability that an
individual has a higher value on one measurement than the other"
(Lakens, 2013, p. 4).
The CLE is therefore the probability that a $Z$-score is greater than a
specific $Z_{\textrm{CLE}}$ (given a standard normal distribution).
For repeated measures the $Z_{\textrm{CLE}}$ is calculated from the means
and standard deviations from two measurements as well as their correlation:
$$
Z_{\textrm{CLE}} =
\frac{M_2 - M_1}{\sqrt{SD_1^2+SD_2^2-2\cdot{}SD_1\cdot{}SD_2\cdot{}r_{12}}}
$$
```{r cle function}
get_cle <- function(mean1, mean2, sd1, sd2, r12) {
  Z_cle <- (mean2 - mean1) / sqrt(sd1^2 + sd2^2 - 2 * sd1 * sd2 * r12)
  Z_cle %>% pnorm(lower.tail = F)
}
```
```{r cle table}
cle_table <- data.frame(format1 = c("Auro", "Auro", "Surround"),
                        format2 = c("Surround", "Stereo", "Stereo"))
cle_table$cle <-
  mapply(function(f1, f2) {
    desc <- descriptives %>% filter(factor == "Transcendence")
    get_cle(mean1 = desc$M[desc$format == f1],
            mean2 = desc$M[desc$format == f2],
            sd1 = desc$SD[desc$format == f1],
            sd2 = desc$SD[desc$format == f2],
            r12 = cor_table$r[cor_table$Formats == paste(f2, f1,
                                                         sep = " -- ")])
  },
  cle_table$format1, cle_table$format2)
cle_table %>% mutate(across(where(is.numeric), round, digits = 4)) %>% 
  knitr::kable(col.names = c("Format 1", "Format 2", "CLE"),
               align = "llc")
```
