---
title: "Reanalysis of a Data set"
author: "Kilian Sander, Yves Wycisk, Reinhard Kopiez"
date: '2022'
output: html_document
csl: inst/apa.csl
---
# Preparation
## Data, Packages, and Design
Load the data:
```{r dataset, messages=FALSE}
load("data/data_raw.Rda")
```
Load relevant packages:
```{r packages, message=FALSE, warning=FALSE, results='hide'}
sapply(c("tidyverse", "MANOVA.RM", "rstatix"), require, character.only = T)
```
Prepare the design with vectors:
```{r formats and GEMS, message=FALSE}
formats <- c("stereo", "surround", "auro")
gems_factors <- c("wonder", "transcendence", "tenderness", "nostalgia",
                  "peacefulness", "power", "joyful_activation", "tension",
                  "sadness")
# item to factor mapping
wonder <- c("moved", "filled_with_wonder", "allured")
transcendence <- c("feeling_of_transcendence", "fascinated", "overwhelmed")
tenderness <- c("affectionate", "tender", "mellowed")
nostalgia <- c("sentimental", "dreamy", "nostalgic")
peacefulness <- c("calm", "serene", "soothed")
power <- c("energetic", "triumphant", "strong")
joyful_activation <- c("joyful", "animated", "bouncy")
tension <- c("agitated", "nervous", "tense")
sadness <- c("sad", "sorrowful", "tearful")
```
## Filtering
Exclude a participant if there are at least two `NA`s for a GEMS factor of one stimulus:
```{r filter cases, message=FALSE}
data_raw$complete <- TRUE
for (p in data_raw$participant) {
  part_data <- data_raw %>% filter(participant == p)
  for (excerpt in c("a", "b")) {
    for (format in formats) {
      for (factor in gems_factors) {
        temp_data <- part_data %>%
          dplyr::select(starts_with(paste(excerpt, format, get(factor), sep = "_")))
        nas <- temp_data %>% t() %>% is.na() %>% sum()
        if (nas > 1) {
          data_raw$complete[p] <- FALSE
        }
      }
    }
  }
}
rm(part_data, temp_data, nas, p, format, excerpt)
data_filtered <- data_raw %>% filter(complete==TRUE)
N <- length(data_filtered$participant)
```
After removing incomplete cases the data set contains `r N` participants.

## Scoring
Calculate mean values for each item per format:
```{r item means, message=FALSE, warning=FALSE}
for (format in formats) {
  for (factor in gems_factors) {
    for (item in get(factor)) {
      item_mean_vector <- data_filtered %>%
        select(ends_with(paste0(format, "_", item))) %>%
        rowMeans(na.rm = TRUE)
      eval(str2expression(paste0("data_filtered$",
                                 format, "_", item,
                                 " <- item_mean_vector")))
    }
  }
}
rm(item, item_mean_vector)
data_filtered <- data_filtered %>% mutate_all(~ifelse(is.nan(.), NA, .))
```
Calculate the mean of each factor per format across its items:
```{r factor means, message=FALSE, warning=FALSE}
for (format in formats) {
  for (factor in gems_factors) {
    factor_mean_vector <- data_filtered %>%
      select(paste0(format, "_", get(factor))) %>% 
      rowMeans(na.rm = TRUE)
    eval(str2expression(paste0("data_filtered$",
                                 format, "_", factor,
                                 " <- factor_mean_vector")))
  }
}
rm(format, factor, factor_mean_vector)
```
## Long format
```{r long formats, message=FALSE, warning=FALSE}
vars_vector <- sapply(formats,
                      function(format) {
                        paste0(format, "_", gems_factors)
                      },
                      USE.NAMES = F) %>% as.vector()
data_long_mult <- data_filtered %>% select(participant, vars_vector)
names(data_long_mult) <- names(data_long_mult) %>%
  gsub("joyful_activation", "joyfulactivation", .)
data_long_mult <- data_long_mult %>%
  pivot_longer(!participant,
               names_to = c("format", ".value"),
               names_pattern = "(.*)_(.*)")
data_long_single <- data_long_mult %>%
  pivot_longer(!c(participant, format),
               values_to = "score",
               names_to = "factor") %>% 
  mutate(factor = recode(factor,
                         joyfulactivation = "joyful activation") %>% tools::toTitleCase(),
         format = format %>% tools::toTitleCase() %>% as_factor())
names(data_long_mult) <- names(data_long_mult) %>%
  gsub("joyfulactivation", "joyful_activation", .)
data_long_mult$format <- data_long_mult$format %>% as_factor()
```
# Analyses
## Descriptives
```{r age summary, include=FALSE}
age_summary <- summary(data_filtered$age)
```
Out of the `r N` participants `r table(data_filtered$gender)[["female"]]`
indicated "female" as their gender whereas `r table(data_filtered$gender)[["male"]]`
indicated "male".
The participants' age ranged from `r age_summary[[1]]` to `r age_summary[[6]]`
with a mean of `r age_summary[[4]]` (*SD* = `r sd(data_filtered$age)`).

```{r descriptives, echo=FALSE, message=FALSE, warning=FALSE}
data_long_single %>% group_by(factor, format) %>%
  summarise(M = mean(score), SD = sd(score)) %>% 
  knitr::kable(col.names = c("GEMS Factor", "Format", "$M$", "$SD$"),
               align = "llcc")
```
```{r mean plot, echo=FALSE, message=FALSE, warning=FALSE}
data_long_single %>% group_by(format, factor) %>%
  summarise(M = mean(score), SD = sd(score)) %>%
  ggplot(., aes(x = factor, y = M, color = format)) +
  geom_point(aes(shape = format)) +
  geom_path(aes(group = format, linetype = format)) +
  ylim(1,5) +
  theme_bw() +
  theme(legend.position = c(.99,.99),
        legend.justification = c(1,1)
        ) +
  xlab("GEMS Factor") +
  ylab("Mean")
```

## Multivariate Analysis of Variance

```{r manova}
#rm_manova <-
#  multRM(formula = cbind(wonder, transcendence, tenderness,
#                         nostalgia, peacefulness, power,
#                         joyful_activation, tension, sadness) ~ format,
#         data = data_long_mult,
#         subject = "participant",
#         within = "format",
#         para = TRUE,
#         iter = 50000,
#         dec = 5,
#         seed = 987)
load(file = "results/rm_manova.Rda")
summary(rm_manova)
```
## Analysis of Variance for *Transcendence*
```{r anova}
transcendence_anova <-
  anova_test(data = data_long_mult,
             dv = transcendence,
             wid = participant,
             within = format,
             type = 3,
             effect.size = "pes",
             detailed = T)
```
## Correlations for *Transcendence*
### Correlations Between the Formats
```{r correlations, include=FALSE}
cor_stereo_surround <-
  cor.test(data_filtered$stereo_transcendence,
           data_filtered$surround_transcendence,
           alternative = "greater",
           method = "pearson",
           conf.level = .95)
cor_stereo_auro <-
  cor.test(data_filtered$stereo_transcendence,
           data_filtered$auro_transcendence,
           alternative = "greater",
           method = "pearson",
           conf.level = .95)
cor_surround_auro <-
  cor.test(data_filtered$surround_transcendence,
           data_filtered$auro_transcendence,
           alternative = "greater",
           method = "pearson",
           conf.level = .95)
cor_table <- data.frame(Formats = c("Stereo -- Surround",
                                    "Stereo -- Auro 3D",
                                    "Surround -- Auro 3D"),
                        r = c(cor_stereo_surround$estimate,
                              cor_stereo_auro$estimate,
                              cor_surround_auro$estimate),
                        p = c(cor_stereo_surround$p.value,
                              cor_stereo_auro$p.value,
                              cor_surround_auro$p.value),
                        LL = c(cor_stereo_surround$conf.int[1],
                              cor_stereo_auro$conf.int[1],
                              cor_surround_auro$conf.int[1]),
                        UL = c(cor_stereo_surround$conf.int[2],
                              cor_stereo_auro$conf.int[2],
                              cor_surround_auro$conf.int[2])) %>% 
  mutate(`Fisher's z` = atanh(r),
         p = sapply(p,
                    function(x) if(x < 10^-3) "< .001"))
```
```{r correlations table, echo=FALSE}
knitr::kable(cor_table,
             col.names = c("Formats", "$r$", "$p$", "LL", "UL", "Fisher's $z$"),
             align = "lccccc")
```

### Average Correlation
To get a mean correlation individual correlation coefficients have to be
Fisher z-transformed.
The Fisher z-transformation is equivalent to the inverse hyperbolic tangent.
```{r average correlations}
avg_cor_z <- mean(cor_table$`Fisher's z`)
avg_cor_r <- tanh(avg_cor_z)
```
Taking the averaged Fisher z-transformed values results in a value of
`r avg_cor_z`.
Back-transformed to Person's *r* the average correlation between the three
audio formats for the GEMS factor Transcendence is `r avg_cor_r`.